---
title: "ST558_HW3_Importing_Data"
author: "Charles Lane"
format: pdf
editor: visual
---

## Task 1 - Conceptual Questions

# 1 - *If your working directory is myfolder/homework/, what relative path would you specify to get the file located at myfolder/MyData.csv?*

  From the R Console, we use the getwd() and setwd() commands to get and set the working directory, respectively. Setting a working directory of a different folder/object with the same root path as the current directory is accomplished with:
  
'setwd("../MyData.csv")'

From the Terminal, commands of simply 'cd ..' will change the directory to a level higher. 'cd ./DIRECTORY' would move the active directory to a location within the current directory.
  
```{r}

```

# 2 - *What are the major benefits of using R Projects?*

  R projects establish a self-contained organizational location. This helps sharing configuration between different parties as all folder or file references can be local.

# 3 - *What is git and what is github?*

  Git is a version control and history software which retains all versions of code and allows for review/comparison of updates to code. Git can therefore be used to collaborate as well. Several different entities may develop on a main branch of code while enabling review by all parties prior to commmitting updates.
  Github is a website or front-end environment for git, including visualization of code projects ("Repos"), and their associated components (branches, descriptive files, etc.)

# 4 - *What are the two main differences between a tibble and a data.frame?*
  A tibble's print results are more organized and display only the top rows/columns.
  Also, as stated in the notes - they do not coerce down to a vector when you subset to only one column using [

# 5 - *Rewrite the following nested function call using baseR’s chaining operator:*

# 6 - *What is meant by long format data and wide format data? Which do we generally prefer for statistical analysis?*

## Task 2 - Reading Delimited Data

### Task 2.1 - Glass Data

First, need to activate the necessary packages for this session.

```{r}
library(tidyverse)
library(readr)
```

Read-in data using the 'readr' package from tidyverse.

```{r}
#reading in data
glass_data <- read_csv(
  "./glass.data",
  col_names = c("ID","RI", "Na", "Mg", "Al", "Si",
                "K", "Ca", "Ba", "Fe", "Type_of_glass"),
  show_col_types = TRUE)
glass_data
```

Now update "Type_of_glass" variable to be a string with descriptive values in place of values of 1-7.
```{r}
glass_data |>
  mutate(
    Type_of_glass = ifelse(
      Type_of_glass == 1,"building_windows_float_processed",
        ifelse(
          Type_of_glass == 2, "building_windows_non_float_processed",
            ifelse(
              Type_of_glass == 3, "vehicle_windows_float_processed",
                ifelse(
                  Type_of_glass == 4,
                    "vehicle_windows_non_float_processed",
                    ifelse(
                      Type_of_glass == 5, "containers",
                        ifelse(
                          Type_of_glass == 6, "tableware",
                            ifelse(
                              Type_of_glass == 7, "headlamps", "ERROR")
                            )
                        )
                    )
                )
            )
          )
      )
```

The preceding chain only updated the 'Type_of_glass' values, but the chain can be extended to include filtering and selection.
```{r}
glass_data |>
  mutate(
    Type_of_glass = ifelse(
      Type_of_glass == 1,"building_windows_float_processed",
        ifelse(
          Type_of_glass == 2, "building_windows_non_float_processed",
            ifelse(
              Type_of_glass == 3, "vehicle_windows_float_processed",
                ifelse(
                  Type_of_glass == 4,
                    "vehicle_windows_non_float_processed",
                    ifelse(
                      Type_of_glass == 5, "containers",
                        ifelse(
                          Type_of_glass == 6, "tableware",
                            ifelse(
                              Type_of_glass == 7, "headlamps", "ERROR")
                            )
                        )
                    )
                )
            )
          )
      ) |>
  filter(
    Fe < 0.2,
    Type_of_glass == "tableware" | Type_of_glass == "headlamps")
```

### Task 2.2 - Yeast Data

Pull in the yeast data
```{r}
yeast_data <- read_fwf(
  "./yeast.data",
  col_positions = fwf_widths(c(12, 6, 6, 6, 6, 6, 6, 6, 6, 3),
                             c("seq_name","mcg", "gvh", "alm", "mit",
                               "erl", "pox", "vac", "nuc", "class")),
  show_col_types = TRUE)
yeast_data
```

Start a chain removing the 'seq_name' and 'nuc' columns.
```{r}
yeast_data |>
  select(-seq_name & -nuc)
```

Continue the chain to determine mean and median for all numeric variables, grouped by 'class'.
```{r}
yeast_data |>
  select(-seq_name & -nuc) |>
  group_by(class) |>
  mutate(across(where(is.numeric), list(mean = mean, median = median), .names = "{.col}_{.fn}"))
```

### Task 2.3 - Combining Excel and Delimited Data

